---
title: "Cleaning USEA text files"
format: html
editor: source
---
Notes to future me
- there are still a couple data sets that need to be converted used claude
- make sure the compiling data still works as expected, some of the data looks a little different with the new datasets




```{r}
library(tidyverse)
library(readr)
library(stringr)
```

Look through all the files in one place to make sure we keep all the data we want and proper column names

```{r, look_at_data}
raw_peek <- function(file_path) {
  # Read with all cols as character so nothing is coerced; header = col names
  df <- read_csv(file_path, col_types = cols(.default = "c"), show_col_types = FALSE)

  list(
    event  = tools::file_path_sans_ext(basename(file_path)),
    header = names(df),
    first  = unname(unlist(df[1, ])))}

peeked <- map(file_list, raw_peek)

max_cols <- map_int(peeked, ~ max(length(.x$header), length(.x$first))) |> max()

pad <- function(x) c(x, rep(NA_character_, max_cols - length(x)))

rows <- map(peeked, function(p) {
  bind_cols(
    tibble(file = p$event, type = c("header", "first_row")),
    setNames(
      as_tibble(rbind(pad(p$header), pad(p$first))),
      paste0("col_", seq_len(max_cols))))})

peek_df <- bind_rows(rows)
```

Loop through and combine all datasets into one

```{r}
read_and_rename <- function(file_path) {
  df <- read_csv(
    file_path,
    col_types = cols(.default = "c"),
    show_col_types = FALSE
  )

  event <- tools::file_path_sans_ext(basename(file_path))

  if (ncol(df) < 10) {
    warning(glue::glue("'{event}' has {ncol(df)} columns (expected 10+). Skipping."))
    return(NULL)
  }

  df <- df[, 1:10]
  names(df) <- target_names

  df |>
    mutate(
      final_score = suppressWarnings(as.numeric(final_score)),
      # rank and all penalty cols kept as character to preserve RF, W, etc.
      event_info  = event
    ) |>
    relocate(event_info)
}

compiled_data <- map(file_list, read_and_rename) |>
  compact() |>
  bind_rows() |>
  add_count(event_info, division, name = "n_entries")
```


Some of the data didn't copy over correctly when I created the CSV files. Check which ones seem to have suspiciously low numbers of entries, and then look at the original files to see if there are any issues with them.
```{r}
event_totals <- compiled_data |>
  count(event_info, name = "total_entries")

event_totals
```


The rankings are all wonky, but we can recalucalted the rankings based on final score.

```{r}
compiled_data_2 <- compiled_data |>
  mutate(
    rank_calc = {
      r <- rep(NA_integer_, n())
      has_score <- !is.na(final_score) & !is.na(suppressWarnings(as.numeric(rank)))
      r[has_score] <- rank(final_score[has_score], ties.method = "min") |> as.integer()
      r
    },
    .by = c(event_info, division)
  )
```

test

```{r}
test <- compiled_data_2 |>
  mutate(diff = as.numeric(rank)-as.numeric(rank_calc) )
```

Final Cleaned dataset and save

```{r}
lev <- c("Intermediate", "Preliminary", "Training",  "Beginner Novice", "Novice",
  "Advanced", "2 Year-old", "3 Year-old", "Yearling", "Modified/Training", 
  "Modified", "Starter", "3-Star", "2-Star", "1-Star", 
  "5 Year-old", "4 Year-old", "4-Star")


final_data <- compiled_data_2 |>
  mutate(
    level = map_chr(division, ~ {
      match <- lev[str_detect(.x, fixed(lev))]
      if (length(match) == 0) NA_character_ else match[1]
    }),
    division2 = map2_chr(division, level, ~ {
      if (is.na(.y)) return(NA_character_)
      .x |>
        str_remove("^\\[HT\\]\\s*") |>
        str_remove(fixed(.y)) |>
        str_remove("^[,\\s]+") |>
        na_if("")
    })
  ) |>
  rename(level_division = division) |>
  rename(division = division2) |>
  mutate(level = ifelse(str_detect(division, "Modified/"), "Modified", level))

```

** double check that the code above still works right with the new data im using for the files that didnt work out the first time

Play around with just training, modified, and prelim
```{r}
tp <- final_data |>
  filter(level %in% c("Training", "Modified/Training", "Modified", "Preliminary")) 

unique(tp$division)
```


E = eliminated ; either too many refusals or jumping in wrong order
W = withdrawn ; before starting course
R = retire ; choose to stop mid course
RRR  = retire ; choose to stop mid course
WW = withdrawn from welfare ; before starting course specific to vet issues


HC = out of competition ; schooling so not for a score or rank
CR = compulsory retirement ; forced to stop mid course due to rule violation (could be because horse looks bad)
TE = technical elimination ; rule issue like jumping a fence not in course or illegal equipment
TETE = technical elimination ; rule issue like jumping a fence not in course or illegal equipment


RF = rider fall ; immediate elimination
RFRF = rider fall ; immediate elimination
MR = mandatory retirement ; similar to CR and is a safety concern
RR = rider requirement ; rider has a safety issue not related to horse


EE = eliminated
EEE = eliminated



```{r}
# codes that indicate unsafe non-completion
safety_codes <- c("E", "EE", "EEE", "R", "RRR", "CR", "RF", "RFRF", "MR", "RR")

# dashes mean horse is already out - treat as NA
dash_pattern <- "^-+$"

tp_analysis <- tp |>
  # exclude HC entries entirely
  filter(
    !xc_jump_pen %in% c("HC") | is.na(xc_jump_pen),
    !sj_jump_pen %in% c("HC") | is.na(sj_jump_pen)
  ) |>
  mutate(
    # extract year for pre/post 2019 split
    year = as.integer(str_extract(event_info, "\\d{4}")),
    period = if_else(year < 2019, "pre_2019", "post_2019"),

    # clean dashes to NA
    xc_jump_clean = if_else(str_detect(xc_jump_pen, dash_pattern), NA_character_, xc_jump_pen),
    sj_jump_clean = if_else(str_detect(sj_jump_pen, dash_pattern), NA_character_, sj_jump_pen),

    # flag safety codes in each phase
    xc_unsafe = xc_jump_clean %in% safety_codes,
    sj_unsafe = sj_jump_clean %in% safety_codes,
    any_unsafe = xc_unsafe | sj_unsafe,

    # three outcomes
    outcome = case_when(
      any_unsafe                                      ~ "unsafe_nonfinish",
      !any_unsafe & !is.na(final_score)               ~ "safe_finish",
      !any_unsafe & is.na(final_score)                ~ "nonsafety_nonfinish"
    )
  ) |>
  # flag horse+rider appearing at more than one level per event
  mutate(
    multi_level_flag = n_distinct(level) > 1,
    .by = c(event_info, horse, rider)
  )

# inspect flagged horse+rider pairs
tp_analysis |>
  filter(multi_level_flag) |>
  select(event_info, horse, rider, level, division) |>
  arrange(event_info, horse, rider)
```


```{r}
tp_analysis |>
  summarise(
    total = n(),
    .by = c(year, level, outcome)
  ) |>
  mutate(
    level   = factor(level, levels = c("Training", "Modified", "Preliminary")),
    outcome = factor(outcome, levels = c("safe_finish", "nonsafety_nonfinish", "unsafe_nonfinish"),
                     labels = c("Safe Finish", "Non-Safety Non-Finish", "Unsafe Non-Finish"))
  ) |>
  ggplot(aes(x = year, y = total, color = level)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ outcome) +
  scale_x_continuous(breaks = seq(2005, 2025, by = 2)) +
  labs(
    x     = "Year",
    y     = "Total Entries",
    color = "Level",
    title = "Outcomes by Year and Level"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```